A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-15 tex/codatadp.tex
more
Minor.

For any $\mu$-value $v$ of the form $\mu x. C( \vec w )$, we may recursively construct a normal value $\nf{v}$
(which we will call the \emph{normal form} of $v$)
that is a renaming of $v$ by replacing all of the bisimilar subterms of $v$ with $x$,
and by replacing all of the non-normal subterms of $v$ with their corresponding normal forms.
For instance, $\nf{(\mu x. \const{C}( \mu y. \const{C}( x ) ))} = \mu x. \const{C}( x )$.

We say that $t \vsimv{x} s$ if and only if $\mu$-terms $t$ and $s$ are syntactically equivalent
for some renaming that avoids capturing variables besides $x$.
For instance, 
$\mu x. \const{D}( x ) \vsimv{y} \mu x. \const{D}( y )$,
$\mu x. \const{C}( x, x ) \vsimv{x} \mu y. \const{C}( x, y )$ and
$\mu x. \const{C}( z, x ) \vsimv{z} \mu y. \const{C}( z, y )$,
but
$\mu x. \const{D}( x ) \not\vsimv{x} \mu x. \const{D}( y )$ and
$\mu x. \const{C}( x, x ) \not\vsimv{y} \mu y. \const{C}( x, y )$.

@@

In the following, we consider only normal models.
Thus, when constructing a model $\M$ for $\Ec$,
it remains only to define how wrong-applied selector terms and variables in $\Ec$ are interpreted.
For the latter, this will be based on the mapping $\Val$ constructed in step two of our calculus.
Given a normal model $\M$, the \emph{completion} of $\Val$, which we will write as $\Val^\ast$, 
is the result of exhaustively applying the following update to $\Val$:

@@

Since this rule removes one variable from the set $FV( \Val )$ and does not add any additional variables to $FV( \Val )$, 
it can only be applied a finite number of times.
We first state the following properties of $\Val$, 
where we write $\muvar( t )$ to denote $t$ if $t$ is a variable, or $x$ if $t$ is of the form $\mu x. u$.
%which amounts to saying that 
%equivalence class are interpreted as values,
%the interpretation of an equivalence class is not sensitive to the context in which it is interpreted,
%and that each equivalence classes has a distinct interpretation.

\begin{lemma} 
\label{lem:model-completion}
The following are true of $\Val^\ast$:
\begin{enumerate}
\item $\nf{(\Val^\ast \ec{x})} \in \M( \tau )$ for all $\ec{x} : \tau$ of $\Ec$,
\item
$t_1 \interp{ t_2 } = \nf{(\Val^\ast \ec{x})}$
for all $t_1\interp{t_2} \in \mathcal{T}(\Val)$ where $\muvar(t_2)=v_{\ec{x}}$,
\item 
$\nf{(\Val^\ast \ec{x})} \vsim \nf{(\Val^\ast \ec{y})}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we must argue that $FV( \Val^\ast ) = \emptyset$, and that $\nf{(\Val^\ast \ec{x})}$ is not cyclic for
$\ec{x} : \tau$ when $\tau \not\in \Codata$.
\rem{TODO}
\qed
\end{proof}

@@

We consider a normal model $\M$ 
that interprets wrongly-applied selectors based on equality information in $\Fc$,
and interprets the variables of $\Fc$ based on the completion $\Val^\ast$ of the mapping $\Val$ from step two of our calculus.
For the latter, let $\M( x )$ be the value in $\M$ that is a renaming of $\nf{(\Val^\ast \ec{x})}$ for each variable $x$ in $\mathcal{T}(\Fc)$.
Notice that due to Lemma~\ref{lem:model-completion}.3, for variables $x, y \in FV( \Fc )$,
\begin{eqnarray}
\label{eqn:m-bij} 
\M( x ) = \M( y ) \text{ if and only if } \ec{x} = \ec{y}
\end{eqnarray}

We first show $\M$ satisfies all equalities in $t_1 \teq t_2 \in \Fc$, where $t_1$ and $t_2$ have type $\tau$.
We show that $\M( t ) = \Val^\ast \ec{t}$ for all terms $t \in \mathcal{T}( \Fc )$, which we show by induction on the size of $t$.

If $t$ is a variable, then $\M( t ) = \Val^\ast \ec{ t }$ by construction.
If $t$ is a constructor term $\const{C}( u_1, \ldots, u_n )$, 
then since \rn{Unify$_2$} does not apply to $\Fc$,
then $\Val^\ast \ec{t}$ must be a term of the form $\mu v_{\ec{t}}. \const{C}( w_1, \ldots, w_n )$,
where say that $\muvar( w_i ) = v_{\ec{y_i}}$ for $i = 1, \ldots, n$.
Since \rn{Unify$_1$} does not apply to $\Fc$, it must be that $\ec{u_i} = \ec{y_i}$ for $i = 1, \ldots, n $.
%Due to Lemma~\ref{lem:model-completion}.2 and by induction
\rem{TODO}

@@

CVC4 was run on each problem several times, with the support for datatypes and
codatatypes either enabled or disabled. The contributions of the cyclicity and
bisimilarity rules were also measured, so find out whether these somewhat
expensive rules are useful in practice. Even when the decision procedure is
disabled, some of the generated axioms typically include basic properties of
constructors and selectors, which the decision procedure would recognize as
tautologies. This allows us to answer the question, \relax{What are the
benefits of activating the decision procedure as opposed to letting
Sledgehammer do what it would normally do?}

>>>>

For any $\mu$-term $t$ of the form $\mu x. C( \vec u )$, we may recursively construct a normal $\mu$-term $\nf{t}$
(which we will call the \emph{normal form} of $t$)
that is a renaming of $v$ by replacing all of the bisimilar subterms of $t$ with $x$,
and by replacing all of the non-normal subterms of $t$ with their corresponding normal forms.
For instance, $\nf{(\mu x. \const{C}( \mu y. \const{C}( x ) ))} = \mu x. \const{C}( x )$.

We say that $t \vsimv{x} s$ if and only if $\mu$-terms $t$ and $s$ are syntactically equivalent
for some renaming that avoids capturing any variable other than $x$.
For instance, 
$\mu x. \const{D}( x ) \vsimv{y} \mu x. \const{D}( y )$,
$\mu x. \const{C}( x, x ) \vsimv{x} \mu y. \const{C}( x, y )$ and
$\mu x. \const{C}( z, x ) \vsimv{z} \mu y. \const{C}( z, y )$,
but
$\mu x. \const{D}( x ) \not\vsimv{x} \mu x. \const{D}( y )$ and
$\mu x. \const{C}( x, x ) \not\vsimv{y} \mu y. \const{C}( x, y )$.

@@

In the following, we consider only normal models.
Thus, when constructing a model $\M$ for $\Ec$,
it remains only to define how wrong-applied selector terms and variables in $\Ec$ are interpreted.
For the latter, this will be based on the mapping $\Val$ constructed in step two of our calculus.
Given a normal model $\M$, we compute the \emph{completion} of $\Val$, which we will write as $\Val^\ast$,
by exhaustively applying the following update to $\Val$:

@@

Since this rule removes one variable from the set $FV( \Val )$ and does not add any additional variables to $FV( \Val )$, 
it can only be applied a finite number of times.
We then replace all values in the range of $\Val^\ast$ by their normal forms.
We first state the following properties of $\Val^\ast$, 
where we write $\muvar( t )$ to denote $t$ if $t$ is a variable, or $x$ if $t$ is of the form $\mu x. u$.
%which amounts to saying that 
%equivalence class are interpreted as values,
%the interpretation of an equivalence class is not sensitive to the context in which it is interpreted,
%and that each equivalence classes has a distinct interpretation.

\begin{lemma} 
\label{lem:model-completion}
The following are true of $\Val^\ast$:
\begin{enumerate}
\item $\Val^\ast \ec{x} \in \M( \tau )$ for all $\ec{x} : \tau$ of $\Ec$,
\item
$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
for all $t_1\interp{t_2} \in \mathcal{T}(\Val)$ where $\muvar(t_2)=v_{\ec{x}}$,
\item 
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we must argue that $FV( \Val^\ast ) = \emptyset$, and that $\Val^\ast \ec{x}$ is not cyclic for
$\ec{x} : \tau$ when $\tau \not\in \Codata$.
\rem{TODO}
\qed
\end{proof}

@@

We consider a normal model $\M$ 
that interprets wrongly-applied selectors based on equality information in $\Fc$,
and interprets the variables of $\Fc$ based on the completion $\Val^\ast$ of the mapping $\Val$ from step two of our calculus.
For the latter, let $\M( x )$ be the value in $\M$ that is a renaming of $\Val^\ast \ec{x}$ for each variable $x$ in $\mathcal{T}(\Fc)$.
Notice that due to Lemma~\ref{lem:model-completion}.3, for variables $x, y \in FV( \Fc )$,
\begin{eqnarray}
\label{eqn:m-bij} 
\M( x ) = \M( y ) \text{ if and only if } \ec{x} = \ec{y}
\end{eqnarray}

We first show $\M$ satisfies all equalities in $t_1 \teq t_2 \in \Fc$, where $t_1$ and $t_2$ have type $\tau$.
We show,
by induction on the size of $t$,
that $\M( t ) \sim \Val^\ast \ec{t}$ for all terms $t \in \mathcal{T}( \Fc )$.
If $t$ is a variable, then $\M( t ) = \Val^\ast \ec{ t }$ by construction.
If $t$ is a constructor term $\const{C}( u_1, \ldots, u_n )$, 
then $\M( t )$ is a renaming of $\nf{(\mu x. \const{C}( \M( u_1 ), \ldots, \M( u_n ) ) )}$ for fresh $x$,
which by the induction hypothesis is bisimilar to $\nf{(\mu x. \const{C}( \Val^\ast \ec{ u_1 }, \ldots, \Val^\ast \ec{ u_n } ) )}$, call this term $t_u$.
Now, since \rn{Unify$_2$} does not apply to $\Fc$,
then $\Val^\ast \ec{t}$ must be a term $t_w$ of the form $\mu v_{\ec{t}}. \const{C}( w_1, \ldots, w_n )$,
that is, an application of the same construction $\const{C}$.
For each $i = 1, \ldots, n$,
say that $\muvar( w_i ) = v_{\ec{y_i}}$.
Due to Lemma~\ref{lem:model-completion}.2, $t_w\interp{ w_i } = \Val^\ast \ec{ y_i }$.
Due to our construction of $\Val^\ast$ and since \rn{Unify$_1$} does not apply to $\Fc$, we have $\ec{u_i} = \ec{y_i}$.

If $t$ is a selector term $s^j_k( u )$,
by the induction hypothesis, $\M( u ) \vsim \Val^\ast \ec{ u }$. 
since \rn{Split} does not apply, 
$\ec{u}$ must contain a term of the form $C_i( s^1_i( u ), \ldots, s^{a_i}_i( u ) )$ for some $i$.

%we show that $w_i$ is bisimilar to the $i^{th}$ argument of
%$\nf{(\mu x. \const{C}( \Val^\ast \ec{ u_1 }, \ldots, \Val^\ast \ec{ u_n } ) )}$.
%In the case that $\mu
%Since $t_c$ is normal

\rem{TODO}

@@

CVC4 was run on each problem several times, with the support for datatypes and
codatatypes either enabled or disabled. The contributions of the cyclicity and
bisimilarity rules were also measured, so find out whether these somewhat
expensive rules are useful in practice. Even when the decision procedure is
disabled, some of the generated axioms typically include basic properties of
constructors and selectors, which the decision procedure would recognize as
tautologies.
%This allows us to answer the question, \relax{What are the
%benefits of activating the decision procedure as opposed to letting
%Sledgehammer do what it would normally do?}
